---
title: 'Lecture 4.2: Inference and Regression'
#subtitle: ""
author: "SIPA U6614 | Instructor: Harold Stolper"
date: 
urlcolor: blue
output: 
  html_document:
    toc: TRUE
    toc_depth: 3
    toc_float: TRUE
    number_sections: TRUE
    highlight: tango
    theme: default
    fig_caption: TRUE
    #fig_width = 7 #Default width (in inches) for figures
    #fig_height = 5 #Default height (in inches) for figures
    #html_preview: TRUE #TRUE to also generate an HTML file for the purpose of locally previewing what the document will look like on GitHub.    
    df_print: tibble # Method to be used for printing data frames. Valid values include "default", "kable", "tibble", and "paged". The "default" method uses print.data.frame. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Load packages:
```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(gapminder)
library(lmtest)
library(sandwich)
library(estimatr)
```

Resources used to create this lesson:

- https://www.econometrics-with-r.org/3-5-cmfdp.html

We're going to explore basic R functionality for inference and regression in the context of gapminder data, this time filtering for only 2007 observations.
```{r, message = FALSE, warning = FALSE}
gap_2007 <- gapminder %>% filter(year == 2007)
```

# Formulas in R

Before we introduce functions for statistical inference and regression analysis, it's useful to introduce the concept of a "formula" in R.

R uses "formulas" to convey the idea of a statistical model, or an unevaluated expression. Formulas are handy for econometrics because they can be used as arguments in other functions to generate special behavior; for example, to point to values of certain variables without evaluating the expression. For example, here is a formula that allows for a linear relationship between some objects `x` and `y`:
```{r}
linear <- y ~ x  #read as: y is a function of x
```

Let's examine how R stores this kind of object:
```{r}
typeof(linear)
attributes(linear)
```

OK, so R refers to this type of object as `language` with `class = formula`. What can we do with this formula? Lots of thing! Like specify a population regression function (PRF), for one:
```{r}
#this is a formula for a linear model
lm1 <- gap_2007$lifeExp ~ gap_2007$gdpPercap
lm1
```
Notice what happens when we print lm1--nothing is evaluating! This allows us to pass the formula for a PRF specifying the relationships between variables as an argument to other functions that can in turn interpret the formula and evaluate in the appropriate context... like functions that perform t-tests and estimate OLS regression, among many other examples.


# Inference

There are many different applications of statistical inference that are relevant for policy research, but the t-test is a something we'll frequently use to test differences in means and regression coefficients, among other applications. 


## `t.test()` with vectors

The `t.test()` function in the `stats` package is a commonly used function for performing one and two sample t-tests on vectors of data (e.g. columns of data in a data frame).

Suppose we want to test the null hypothesis that the difference in mean life expectancy across Asian countries differs from European countries. We need to specify the two columns of data to use for computing each group mean. The life expectancy data doesn't exist in separate columns by continent, but we can easily create them by filtering the gapminder data and assigning to new continent-specific data frames.
```{r, warning = FALSE}
#create new data frames filtered by continent
gap_2007_asia <- gap_2007 %>% filter(continent == "Asia")
gap_2007_eur <- gap_2007 %>% filter(continent == "Europe")

#conduct t-test for diff in mean lifeExp by referencing appropriate lifeExp columns
t.test(gap_2007_asia$lifeExp, gap_2007_eur$lifeExp, var.equal = FALSE)

#store results of t-test as its own object to refer back to statistics
ttest1 <- t.test(gap_2007_asia$lifeExp, gap_2007_eur$lifeExp, var.equal = FALSE)

#check the list of stored objects from this test that you can refer back to
str(ttest1)

#get t-stat, handy for in-line references in your R Markdown doc
round(ttest1$statistic, 2)

#get associated p-value
round(ttest1$p.value, 2)
```
The results indicate a t-stat of `r round(ttest1$statistic, 2)` and associated p-value of `r round(ttest1$p.value, 2)`, with a difference in mean life expectancy of almost 7 years that is statistically significant at the 1% level.

But wait a minute, **what formula do we want R to use to estimate the standard error (SE) of the difference??** 

We generally want to allow for unequal variances in the different group populations, which is equivalent to robust SEs in the analogous bivariate regression setting (we'll get to the bivariate regression framing shortly). Here the default option is to assume the sample means are drawn from populations with unequal variances, but can we still explicitly specify the variance formula as an additional option. (Consult Stock and Watson Chapter 3.4 (3rd edition) for a refresher on difference-in-mean tests and the appropriate SE formula, as well as the Quant II review materials posted to the [course website](https://hreplots.github.io/U6614/).)

Notice how the SE of the difference in means changes slightly when the SE formula does NOT allow for unequal variances:
```{r, message = FALSE, warning = FALSE}
#conduct t-test allowing for unequal population variances
t.test(gap_2007_asia$lifeExp, gap_2007_eur$lifeExp, var.equal = TRUE)
```
We can also specify other function options to specify the t-test interest we're interested in if we don't want to rely on the default options: 

- `conf.level = 0.99` specifies 99% confidence
- `mu = 10` specifies a null hypothesized value other than zero, if appropriate
- `alternative = "less"` specifies the appropriate one-tailed test if a two-tailed test isn't appropriate


## `t.test()` with formulas

An alternative approach is to pass a single formula as an argument to the `t.test()`, instead of two vectors of data. Before we see this in action, first let's (re)introduce a function from another package (`dummy_cols()` in the `fastDummies` package) that creates separate dummy variables for each category of a categorical variable (recall we used in class during week 3 with the subway fare evasion microdata). In this, case let's use this function to create dummy variables indicating each continent (e.g. continent_Asia).
```{r, message = FALSE, warning = FALSE}
#create dummy variables for each category of continent (as new columns)
gap_2007_d1 <- fastDummies::dummy_cols(gap_2007, select_columns = "continent")
str(gap_2007_d1)
```
Note that the syntax `fastDummies::dummy_cols()` is a way of accessing the `dummy_cols()` function in the `fastDummies` function without loading the full package. This known as a *lazy load* of a package.

Now let's run the same t-test as above but using a formula as the argument specifying the null hypothesis to test, **H~0~: mean life expectancy across Asian countries differs from European countries**. 

The key idea is to specify the formula for a bivariate regression that yields an equivalent t-statistic: 

*lifeExp~i~ = b~0~ + b~1~continent_Asia~i~ + u~i~ , *

estimated over Asian and European countries only. We just need to pass this formula to the `t.test()` function, along with another argument specifying the data source in which to look for elements references in the formula. (If you're fuzzy on the equivalence between a conventional difference-in-means t-tests and the bivariate regression approach, please consult [Video 2.2.a Dummy Variables and Regression Analysis](https://drive.google.com/file/d/1-kM9FlwZz7Kx1ynbe3_9ArhHtfHNGNz6/view?usp=sharing) from Quant II.)

```{r, message = FALSE, warning = FALSE}
#filter gap_2007 to exclude countries outside of Asia and Europe
gap_2007_asia_eur <- gap_2007_d1 %>% filter(continent == "Asia" | continent == "Europe")

#conduct t-test for diff in mean lifeExp by passing formula for appropriate bivariate PRF
t.test(lifeExp ~ continent_Asia, data = gap_2007_asia_eur, var.equal = FALSE)
```
Notice we get the same t-statistic as before, `r round(ttest1$statistic, 2)`!

Both syntactic approaches to the `t.test()` will come in handy, depending on how your data is set up and the particular null hypothesis you want to test.


# OLS Regression

We've already seen how formulas can be used to specify a population regression function (PRF). All we need to do is pass the formula to a function that will estimate the corresponding OLS regression. Enter the `lm()` function for fitting linear models, along with `summary()` function to report results.
```{r, message = FALSE, warning = FALSE}
#estimate bivariate linear regression of lifeExp on continent_Asia (dummy variable)
ols1 <- lm(lifeExp ~ continent_Asia, data = gap_2007_asia_eur)
summary(ols1) # view OLS estimates

#we can call str(ols1) to view all elements stored in the ols1 object
#let's pull out the coefficient of interest, beta1_hat
ols1$coefficients[2] 
```

The results look pretty good, the coefficient on continent_Asia gives us the same difference in mean life expectancy between Asian and European countries as with `t.test()`... except we haven't specified robust SEs, which means we haven't allowed for unequal variances in the underlying population of life expectancy between countries in the two continents. 

There is one more step to go to perform inference on regression coefficients using robust SEs... we need to install and load two more packages: `lmtest` and `sandwich`. The former allows us to call the `coeftest()` function to conduct inference on estimated coefficients, and the latter allows us to specify the appropriate covariance matrix used to estimate robust SEs.
```{r, message = FALSE, warning = FALSE}
#conduct inference on coefficient from linear model ols1
lm1 <- coeftest(ols1, vcov = vcovHC(ols1, type="HC1")) # type="HC1" specified robust SE formula
lm1

#the results from coeftest are generally better to display, and we can easily pull out the robust SE for beta1_hat
lm1[2,2] #check to str(lm1) to confirm why this is the right syntax
```
Now we have the same t-statistic (`r round(lm1[2,3],2)`) as with the conventional difference-in-means t-test assuming unequal population variances.


## Model specification 

We've already seen how `fastDummies::dummy_cols()` is a useful function for creating dummy variables that can be in included as variables in a regression. Some other commonly used tools for specifying non-linear models include:

- log transformations: `log1p(x)`(uses a particular form of the natural log function)
- interactions: `x * y`
- quadratic terms: `I(x^2)` (uses the `I()` or "as-is" operator)

Consider the following non-linear PRF:

*lifeExp~i~ = b~0~ + b~1~pop~i~ + b~2~continent_Asia~i~ +b~3~popXcontinent_Asia~i~ + b~4~gdpPercap~i~ + + b~5~gdpPercap^2^~i~ + u~i~ *

We have no good reason to estimate this particular model so let's not dwell on the results, it's just a hypothetical PRF that puts (most of) the above specification tools to use. Here is the R syntax for estimating this PRF:
```{r}
ols2 <- lm(lifeExp ~ pop * continent_Asia + gdpPercap + I(gdpPercap^2), data = gap_2007_asia_eur)
summary(ols2)
```


## Plotting sample regression functions

Another use of linear models is plotting regression estimates (the sample regression function) under `ggplot` using the `geom_smooth()` geometry. We'll spend more time on visualizations with ggplot in a few weeks, but for now here is an example:

```{r}
#add linear prediction line to scatter plot
ggplot(data = gap_2007,                               #specify data 
       aes(x = log1p(gdpPercap), y = lifeExp)) +      #set aesthetic mapping
  geom_point() +                                      #add geometry layer for scatterplot
  geom_smooth(method = 'lm_robust', formula = y ~ x)  #add geometry layer for regression line using lm_robust from the estimatr package
```

The only catch here is that standard linear model geometry, `method = 'lm'`, does not use robust SEs. A solution is to specify `method = 'lm_robust'` that requires the `estimatr` package. If you're interested, [this](https://stackoverflow.com/questions/45313482/ggplot2-how-to-get-robust-confidence-interval-for-predictions-in-geom-smooth) Stackoverflow post explains this issue a bit more.
