---
title: "Your title here"
author: Your name and UNIs here
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  pdf_document:
    keep_tex: false
    fig_caption: true
    latex_engine: pdflatex
header-includes:
  -  \usepackage{hyperref}
biblio-style: apsr
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
# spacing: double
endnote: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE,
                      message=FALSE, warning=FALSE, echo=FALSE,
                      fig.path='figs/',
                      cache.path = '_cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })
```


```{r, libraries}
library(tidyverse)
library(stargazer)
```

```{r}
load("datasets/vf_joined.RData")
vf_joined <- vf_joined %>% 
  mutate(vfund_allocation = vfund_allocation / 1000, 
         vfund_per_poor = vfund_per_poor / 1000)

load("datasets/zip_cross.rda")
load("datasets/zip_panel.rda")
load("datasets/ym_race.rda")
```

# 1. Introduction

`Academic workflow` is at a crossroads. The *American Journal of Political Science* (*AJPS*) announced a (my words) ["show your work" initiative](http://ajps.org/2015/03/26/the-ajps-replication-policy-innovations-and-revisions/) in which authors who are tentatively accepted for publication at the journal must hand over the raw code and data that produced the results shown in the manuscript. The editorial team at *AJPS* then reproduces the code from the manuscript.

Clearly state your research questions. 

## 1.1 Research questions

This paper explores the following research questions in the context of ABC with XYZ data:

\begin{itemize}
\item Is R better than Stata?
\item Is U6614 the easiest course in Columbia?
\end{itemize}


## 1.2 Motivation and policy context

Motivate your study and describe the policy context.



# 2. Background

More policy background details here (if necessary, otherwise fold into Section 1).


# 3. Data description

Describe data sources, representative population, and any other context the reader should know about the data as it pertains to the analysis you do. More details can be included in a Data Appendix.



# 4. Descriptive statistics

Describe the distribution of the key variables you’re analyzing. Focus mostly on your treatment or policy variable(s) of interest and how they vary across relevant groups and/or time, as well as key covariates that need to be account for.

For example, you should include difference-in-means tables or plots, and/or time series plots of key variables (by subgroups, when appropriate).


```{r results = "asis"}
stargazer(as.data.frame(vf_joined), type = "latex", 
          summary.stat = c("n", "mean", "sd", "min", "max"), #specify columns
          covariate.labels = c("VF allocation", 
                               "Population in poverty (1000s)", 
                               "VF allocation/poor residents"),
          digits = 2, column.sep.width = "15pt", #other formatting
          notes = "Village Fund amounts in millions of rupiah.",
          header = FALSE, title = "Put your title here")
```

This is a good table. You can add notes for more documentation of technical details.


```{r, fig.cap="Put your title here", fig.width=7, fig.height=4}

ggplot(data = ym_race, 
       aes(x = date, y = si_1000, group = black75)) + 
  geom_line(aes(color = black75)) +
  labs(x = "Date", y = "Shutoff rate per 1,000 residents",
       color = 'Census tract racial composition')
```

Make sure X and Y are clearly labeled and the units are clear. 


# 5. Empirical strategy

Carefully describe the econometric methods you use and main regression specifications using clear and unambiguous notation.

Please use Piazza posts to share clarifying questions with your classmates and instructors.


# 6. Findings

Can be split into multiple sections if appropriate.


# 7. Conclusion

Summarize your key findings and policy implications of these findings. Discuss the limitations of your analysis and next steps.


\newpage

# 8. References

\newpage

# 9. Appendices

Use appendices for more detailed data description, and supplementary tables or charts that provide supporting information that is useful but not central for the story you are telling with your data.

