---
title: "Formatted Output & R Markdown Presentations"
output:
  ioslides_presentation: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{css, echo = FALSE}

table, td, th {
  border: none;
  padding-left: 1em;
  padding-right: 1em;
  margin-left: auto;
  margin-right: auto;
  margin-top: 1em;
  margin-bottom: 1em;
}

```

# Creating presentations with R Markdown

## Brief overview of ioslides | Creating slides from R Markdown

1. To create an **ioslides** presentation from R Markdown, specify *ioslides_presentation* in the YAML header (**beamer** and **slidy** are other R Markdown presentation types)

2. Use # for a new section; and ## for a new slide. Use | for adding a subtitle

3. Can set slides to display smaller text {.smaller} or display the slide content incrementally {.build}, for example:

```{r, echo = TRUE}
## Slide title {.smaller .build}
```

[Here](https://garrettgman.github.io/rmarkdown/ioslides_presentation_format.html) is a short tutorial. See Chapter 4.1 in [R Markdown: The Definitive Guide]( https://bookdown.org/yihui/rmarkdown/ioslides-presentation.html) for a more comprehensive treatment.


# Including output in your slides

## Loading packages {.build}

This presentation uses the following packages to generate and display output:

<div>
- *tidyverse*
<div>
- *stargazer*
<div>
- *kableExtra* 
<div>
- *knitr* 
<div>
- *plm*
<div>
- *lmtest* 
<div>
- *multiwayvcov*

```{r}
library(tidyverse)
library(stargazer)
library(kableExtra)
library(knitr)
library(plm)
library(lmtest)
library(multiwayvcov)
```

## Loading data for today's examples {.smaller}

Load `vf_joined` dataframe from strings and dates lecture
```{r}
load("formatted_output_slides_data.RData")
head(vf_joined)
```

Load Detroit data (cross-sectional and panel; cross-sectional shown here)
```{r}
head(zip_cross)
```


## Difference-in-means table 

Suppose we're interested in comparing the *poverty rate* and *vfund_per_poor* of regencies between Java and Sulawesi islands.

[Here](https://sebastiansauer.github.io/multiple-t-tests-with-dplyr/) is a tutorial for making this kind of table, shown below (this example does not rely on the stargazer package). 

```{r preprocessing 1}
vf_province <- vf_joined %>%
  mutate(island = if_else(str_starts(province, 'Jawa') | 
                            province == 'Banten' | 
                            province == 'Di Yogyakarta', 
                          'Java', 
                          if_else(str_starts(province, 'Sulawesi') | 
                                    province == 'Gorontalo',
                                  'Sulawesi', 'Other'))) %>%
  filter(island != 'Other') %>%
  select('poverty', 'vfund_per_poor', 'island')
```

```{r preprocessing 2}
vf_diff <- vf_province %>%
  gather(key = variable, value = value, -island) %>%
  group_by(island, variable) %>%
  summarise(value = list(value)) %>%
  spread(island, value) %>%
  group_by(variable) %>%
  mutate(mean_java = round(mean(unlist(Java)), 2),
         mean_sulawesi = round(mean(unlist(Sulawesi)), 2),
         diff = round(mean_java - mean_sulawesi,2),
         p_value = t.test(unlist(Java), unlist(Sulawesi), var.equal = FALSE)$p.value) %>%
  mutate(p_value = sprintf("%.3f", round(p_value, 3))) %>%
  select(-Java, -Sulawesi)

vf_diff$variable <- c("Population in poverty (1000s)", "VF allocation/poor resident")
```

```{r output}
kable(vf_diff, 
      align = "l",
      col.names = c("Variable", 
                    "Java", 
                    "Sulawesi", 
                    "Difference", 
                    "p-value")) %>% 
      kable_paper("hover", 
                  full_width = F, 
                  html_font = "Helvetica") %>% 
      kable_styling(font_size = 20)
```

## Using Stargazer for formatted output {.smaller}

`stargazer` is an R package that creates LATEX code, HTML code and ASCII text for well-formatted regression tables, with multiple models side-by-side, as well as for summary statistics tables, data frames, vectors and matrices."

[Here](https://www.jakeruss.com/cheatsheets/stargazer) is a good tutorial with more stargazer examples and formatting options.


## Regression tables {.smaller}

Columns 1 and 2 show results from the cross-sectional regression outputs without and with vacancy rate as a control variable, respectively.

```{r results = 'asis'}
se_robust <- function(x) 
  coeftest(x, vcov = vcovHC(x, type = "HC1"))[,2]

cross_total_1 <- lm(total_obs_1000 ~ si_1000, 
                      data = zip_cross,
                      weight = pop2015)

cross_total_2 <- lm(total_obs_1000 ~ si_1000 + vac_res_p100, 
                      data = zip_cross,
                      weight = pop2015)

mdls <- list(cross_total_1, cross_total_2)

se <- list(se_robust(cross_total_1), se_robust(cross_total_2))

stargazer(mdls, align = TRUE, type = "html", 
          keep.stat = c("n", "adj.rsq"), #specify regression statistics to keep/omit
          se = se, #specify SEs to display based on the list we created
          notes = "Robust SEs shown in parentheses.", notes.align = "l")
```


## Regression tables {.smaller}
Full regression equation:

$$total\_obs\_1000_{zt} = \beta_0 + \beta_1 si\_1000_{zt} + \beta_2 vac\_res\_p100_{zt} + \phi_z + \theta_t + u_{zt}$$

Columns 1 and 2 show results from the fixed effects regressions we estimated in class without the vacancy rate control, columns 3 and 4 include vacancy rate as a control variable.

```{r results = 'asis'}
se_robust <- function(x) 
  coeftest(x, vcov = vcovHC(x, type = "HC1"))[,2]

se_cluster <- function(x)
  coeftest(x, cluster.vcov(x, cbind(zip_panel$zip5),
                                 df_correction = T))[,2]

panel_total_1a <- lm(total_obs_1000 ~ si_1000 + as.factor(zip5) + as.factor(ym), 
                        data = zip_panel, 
                        weight = pop2015)

panel_total_1b <- panel_total_1a

panel_total_2a <- lm(total_obs_1000 ~ si_1000 + vac_res_p100 + as.factor(zip5) + as.factor(ym),
                      data = zip_panel, 
                      weight = pop2015)

panel_total_2b <- panel_total_2a

mdls <- list(panel_total_1a, panel_total_1b, panel_total_2a, panel_total_2b)

se <- list(se_robust(panel_total_1a), se_cluster(panel_total_1b),
           se_robust(panel_total_2a), se_cluster(panel_total_2b))

mean_y <- round(mean(zip_panel$total_obs_1000), 2)

stargazer(mdls, align = TRUE, type = "html",
          omit=c("zip5", "ym"), #omit coefficients for FE terms
          se = se, #specify SEs to display based on the list we created
          keep.stat = c("n", "adj.rsq"),
          add.lines = list(c("Zip code FE", rep("Yes", 4)),
                           c("Month FE", rep("Yes", 4)),
                           c("Clustered SEs", "No", "Zipcode", "No", "Zipcode"),
                           c("Mean dep var", rep(mean_y, 4))),
          notes = "Robust standard errors are shown in parentheses unless otherwise indicated.",
          notes.align = "l")
```

# Additional ioslides formatting


## Creating bullet points {.build}

<div>
- This is a bullet point
</div>
<div>
- This is another bullet point
</div>
<div>
- This is another bullet point
</div>


## Add a plot

```{r}
load("datasets/ym_race.rda")

ggplot(data = ym_race, 
       aes(x = date, y = si_1000, group = black75)) + 
  geom_line(aes(color = black75)) +
  labs(x = "Date", y = "Shutoff rate per 1,000 residents",
       color = 'Census tract racial composition')
```

Put your explanation here.
