---
title: 'Lecture 3.1: Data Analysis Tools'
author: 'SIPA U6614 | Instructor: Harold Stolper'
date: 
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: yes
    highlight: tango
    theme: default
    fig_caption: yes
    df_print: tibble
  pdf_document:
    toc: yes
    toc_depth: '3'
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Load packages:
```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(gapminder)
```

Resources used to create this lesson:

- https://stat545.com/factors-boss.html
- https://stat545.com/dplyr-single.html#group_by-is-a-mighty-weapon
- https://mgimond.github.io/ES218/Week04c.html
- https://r4ds.had.co.nz/data-visualisation.html


# Importing data

The `utils` package that is automatically loaded with R includes a set of functions for importing data (e.g. `read.csv()`). Comparable import functions included in the `readr` package (e.g. `read_csv()`) are generally faster and import data as tibbles that work well with `tidyverse` packages. Here we'll focus on functions for importing data in the `readr` package , which is included as part of the [Tidyverse](https://www.tidyverse.org/packages/) suite of packages.


## Comma delimited files (.csv) 

Many datasets are stored at text files with rows of data and values for each column separated by commas (or some other *delimiter*, such as a semi-colon or tab). You can easily import csv files using the `read_csv()` function, as in the below code chunk. (Note that the equivalent function from the `utils` package is `read.csv()`).

```{r eval = FALSE}
df1 <- read_csv("yourfilename.csv")
```

The default is for strings to be stored as characters rather than factors (unlike `read.csv()` in the `utils` package). One common option you can set is to specify which values are read in as `NA`, in this example we'll treat blank character values as NAs. Another `read_csv()` option allows you to specify the column names assigned to the imported data.
```{r eval = FALSE}
varnames <- c("c1", "c2", "c3", "c4)
df1 <- read_csv("yourfilename.csv", na = "", col_names = varnames)
```

`read_tsv()` is a similar function specifically for reading in tab-delimited text files of data. `read_csv()` and `read_tsv()` are known as wrapper functions. Wrapper functions call another function that does the real work, but "wrap" that function in way that allows for simpler syntax and is often easier to work with. Wrappers often specify some of the common options for you, but there are often many more options you can explore in the documentation.

In this case it's `read_delim()` that does the real work when you call `read_csv()`, `read_csv()` simply tells `read_delim()` to specify the options to treat commas as the file delimiter.

`fread()` in the `data.tables` package is another powerful tool for effortlessly reading in data without having to specify many (or any!) options.


## Excel files

An Excel file is effectively a set of data tables, with each worksheet as its own data table. The `readxl` package provides a function for recognizing sheet names (`excel_sheets()`) and another function for importing specific sheets in a workbook of data (`read_excel()`). You can use `excel_sheets()` to obtain worksheet names and then tell R to read those worksheets by specifying them as arguments when calling `read_excel()`. You can also simply refer to a worksheet by its index number in the workbook, as in the below code chunk. Just make sure you remember to install and load the `readxl` package first.
```{r eval = FALSE}
library(readxl)
excel_sheets("yourworkbook.xls") #check worksheet names
df2 <- read_excel("yourworkbook.xlsx", sheet = 2) #import 2nd worksheet
```

## R data files

R also has its own data file format, .rds, that you can use to save objects (data frames as well as other R objects) and easily load them later using the base R `readRDS()` function.
```{r eval = FALSE}
#save an object to a file
  saveRDS(df2, file = "mydataobject.rds") #here df2 is a data frame in memory

#load the rds file with the data object df2 back into RStudio and store as df3
  df3 <- readRDS("mydataobject.rds")
```


## Fixed width files

A fixed width file is a text file where each column has a maximum number of characters (i.e. a "fixed width"). There are no delimiters used within rows to separate values for each column. This means you have tell R, or any statistical software package, which characters should be assigned to which variables. `read_fwf()` is a `readr` package that makes it relatively easy to read in fixed width files.


# Exporting data

There may be instances where you want to export data for use in other applications, such as mapping software or web-based applications. Also included in the `readr` package, the `write_csv()` function is a wrapper function for exporting .csv files. Unless you specify a different directory, `write_csv()` will write to the working directory.
```{r eval = FALSE}
write_csv(yourdataframe, "yourfilename.csv")
```


# Appending data sets

Appending is simply stacking rows of data on top of each other. For example, you may have two identically formatted data frames, each with data for a different year, that you want to combine into a single data frame spanning both years. 

The most important requirement for effective appending is to make sure all of your datasets have comparable information stored under the same column names, and with values coded in a consistent manner. For example, if you appended individual-level data from one year with height measured as inches to another year of data where height is measured as centimeters, you'd end up with a useless set of height measurements. You'd first want to convert height in one dataset to the same unit of measurement as the other, then you'd want to make sure height is stored under the same column name in both datasets.

Here is the generic syntax for appending using `rbind()`, the base R function for binding rows:
```{r eval = FALSE}
df_combined <- rbind(df1, df2) #append two data frames, df1 and df2
```

`rbind()` only works if both data frames have the exact same columns. If you have an additional column of data in one data frame that you want to keep when you append them together (assigning NA values for this column to observations from the other set), try `rbind.fill()` in the `plyr` package.

Combining **columns of data** is known as a **join**, which we'll explore next week.


# Inspecting columns of data

We've already seen that `str()` is a very handy tool for inspecting the structure of different data objects. Before we start cleaning and summarizing data, let's review some basic functions for inspecting data frames and columns of data. Let's focus on the familiar gapminder dataset and the `continent` column as an example.

`str(gapminder)` shows us the structure of the gapminder data frame (loaded automatically with the gapminder package) including the dimensions (columns x rows), column names, data type/class, and snapshot of values for each column.
```{r}
str(gapminder)
```

`head()` and `tail()` show the first or last parts of an object (an optional argument allows you to specify how many observations to show). This is useful for observing high/low values of a variable after calling `arrange()`.
```{r}
#display the 5 gapminder countries with the highest life expectancy in 2007
gapminder %>% 
  filter(year == 2007) %>% 
  arrange(desc(lifeExp)) %>% 
  head(n= 5)
```

Remember that if you pass a data frame to a tidyverse function through a pipe, you don't need to refer to the data frame as an argument in the next function after the pipe operator. 
```{r}
#piping is super useful for processing data
  gapminder %>% head(n = 3)

#you can also call tidyverse functions without pipes, this is ok for a single function call
  head(gapminder, n = 3)
```

Within a pipeline, you can also just refer to columns by name when specifying function arguments (when applicable). But if you call a base R function that isn't part of the tidyverse package, you'll need to use the `$` syntax to access specific columns within a data frame. Here are a few base R functions that help us inspect specific columns of data (i.e. variables).
```{r}
typeof(gapminder$continent)
attributes(gapminder$continent)
class(gapminder$continent)
levels(gapminder$continent)
nlevels(gapminder$continent)

```


```{r eval = FALSE}
#if you try to run the next line, it won't work
  gapminder %>% nlevels(continent) #this will generate an error!

#why not? nlevels() isn't a tidyverse function so it won't accept arguments through a pipe in the same way
```

# Basic data cleaning tools

The `dplyr` functions introduced last week are useful tools for understanding and managing your dataset: `filter()`, `arrange()`, `select()`, `mutate()` and `rename()`. They help us understand what columns of data look like... and may suggest certain data cleaning tasks to make the information we want to work with clear, consistent, and ready for statistical analysis.

Here are some basic tools for recoding information stored in columns of data.

## Recoding values

`recode()` is another dplyr function for replacing numeric, character or factor values. Say we want to recode gapminder observations listed with the country name "West Bank and Gaza" to instead match up with the name used by the UN and EU, the "Occupied Palestinian Territories". We can use the `recode()` function as part of the argument for the `mutate()` function in order to create a new variable called `country_clean`.
```{r}
#recode West Bank and Gaza observations to Occupied Palestinian Territories
g2 <- gapminder %>% 
  mutate(country_clean = recode(country, "West Bank and Gaza" = "Occupied Palestinian Territories")) 

#let's check if it worked by printing observations where country == "West Bank and Gaza"
g2 %>% filter(country == "West Bank and Gaza") 
```
Note that `country` column is a factor, so categorical values (levels) are stored as integers but displayed as characters. We can refer to the levels (category labels) themselves when using `recode()`.

`recode()` can also be used to recode values in numeric, character, and logical columns of data.

`replace()` is a base R function that can also be used for recoding values.


## Setting factor levels

Suppose that we have a factor variable `var1` with 4 categories, but we want to exclude observations in the fourth category as `NA` because we're unclear what this category means. One way to do that is to use `factor()` to explicitly set the levels to use, which in turn treats all excluded factor levels as `NA`.

```{r eval = FALSE}
#create a vector including the desired factor levels to use for var1
levels_to_keep <- c("cat1", "cat2", "cat3")

#use the contents of this vector to set the levels attribute of column var1 in data frame df1
df1 %>% mutate(var1.fac = factor(var1, levels = levels_to_keep))
```

## Re-ordering factor levels

The `forcats` package included in the `tidyverse` includes a number of tools for working with factors. This includes a set of functions for re-ordering factor levels, which can be useful when we summarize factor variables.
```{r message = FALSE, warning = FALSE}
#the default level is alphabetical
gapminder$continent %>% levels()

#reorder levels in reverse alphabetical order
gapminder$continent %>% 
  fct_rev() %>% 
  levels()

#reorder levels by frequency 
fct_count(gapminder$continent)
gapminder$continent %>% 
  fct_infreq() %>% 
  levels()

#reorder levels by putting Europe first
gapminder$continent %>% 
  fct_relevel("Europe") %>% 
  levels()

#reorder levels based on minimum lifeExp among all observations in each continent
gapminder$continent %>% 
  fct_reorder(gapminder$lifeExp, min) %>% 
  levels()
```
You can take a look at the forcats [documentation](https://forcats.tidyverse.org/) for a cheat sheet that describes other functions for working with factor variables.


## if_else()

`if_else` is a useful dplyr function for manipulating data in (non-factor) columns of data based on conditional statements. To get the gist of how `if_else` works, let's create a binary (logical) variable called `cont_asia` equal to `TRUE` if `continent == "Asia"` and equal to `FALSE` for other continents.
```{r message = FALSE, warning = FALSE}
g3 <- gapminder %>% 
  mutate(cont_asia = if_else(continent == "Asia", TRUE, FALSE) ) 
str(g3)
```
We'll check to make sure this mutate worked correctly in the next section. 

Also note that base R includes its own `ifelse` function that works a bit differently than the dplyr `if_else` function. The dplyr version is more rigid when it comes to working with different data types.

`mutate_if()` is another tool for creating variables in ways that depend on certain specified conditions.


# Summary statistics

**Summary statistics**, or **descriptive statistics**, are terms we'll use to describe information about the distribution of a random variable. Before we start exploring relationships between variables, we generally want to understand the distribution of individual variables. Some examples of summary statistics that tell us something about the distribution a single variable:

- mean, median, min, max, and standard deviation are statistics that characterize the distribution of a *continuous variable*
- frequency tables characterize the distribution of *categorical variables* 

Visualizing descriptive statistics can often be more digestible and informative, we'll get there in shortly!


## summary()

`summary()` is a base R function that summarizes objects in ways that depend on the class of the object. It can be a handy way to quickly generate summary statistics for columns of data, though the display format varies depending on the class of the argument. Here are some examples.

```{r message = FALSE, warning = FALSE}
summary(g3)

#note how summary treats factors, characters, and logical data types differently
summary(g3$continent) #summary of factor data
summary(as.character(g3$cont_asia)) #summary of character data
summary(g3$cont_asia) #summary of logical data
```


## ()

A more flexible function for computing summary statistics is the dplyr `()` function. `summarise()` stores results as data frames (as tibbles, as is the norm in the Tidyverse) and can be used to perform more complicated calculations in combination with `group_by()`. 

This can be helpful for characterizing *conditional distributions*, for example computing statistics for one variable conditional on the value of another variable. (Crosstabs, discussed in the next section, are another tool for showing *conditional* or *joint* distributions.)

```{r message = FALSE, warning = FALSE}
#display the mean life expectancy for 2007 (and the number of observations)
gapminder %>%
  filter(year == 2007) %>%  
  summarise(n_countries = n(), avg_lifeExp_2007 = mean(lifeExp))
```

Here is a list of "helper" functions we can use within `summarise()` (also accessible via `?summarise`):

| Function | Description |
|----------|-------------|
| `n`  |    count   |
| `n_distinct`  |   count unique values    |
| `mean`  | mean      |
| `median`  |   median    |
| `max` | largest value |
| `min` | smallest value |
| `sd`  | standard deviation |
| `sum`  |  sum of values    |
| `first` | first value |
| `last` | last value |
| `nth`  |  nth value     |
| `any`  |  condition true for at least one value |


## Frequency tables and crosstabs

The base R `table()` function is a quick tool for generating frequency tables (also known as one-way tables) and crosstabs (two-way tables). Note that passing the factor column `continent` as an argument to the `table()` function yields equivalent results as passing it to the `summary` function.
```{r}
table(g3$continent)
```

Let's look at a crosstab between `continent` and `cont_asia` to make sure we correctly created the logical variable indicating Asian observations. We can also look at column or row proportions (conditional distributions) instead of cell counts.
```{r message = FALSE, warning = FALSE}
#joint distribution
table(g3$continent, g3$cont_asia)

#distribution of continent conditional on being in Asia (admittedly not very interesting)
prop.table(table(g3$continent, g3$cont_asia) , 2) %>% round(2)
```

Note that there may be other packages with functions specifically designed for the types of data you're working with. For example, `fct_count()` in the `forcats` package shows a frequency table for factor variables, yielding equivalent results to the base R `table()` function but with nicer presentation.
```{r message = FALSE, warning = FALSE}
fct_count(g3$continent)
```


# Grouping

The `group_by()` function allows us to add extra structure to your data by telling R to recognize groups. This allows for calculating statistics at the group-level using `summarise()` and `mutate()`, both of which respect groups.

`summarise()` will generally take a dataset and return a single observation (per group) with specified summary statistics, such as counting observations or computing means (in each group, if groups are specified).  

Combined with `mutate()`, `summarise()` and other dplyr functions, `group_by()` enables exploratory data analysis tools that we will use throughout this course. Let's start by using these functions with the gapminder data frame to count observations for certain groups of country-year observations and make some calculations at the continent level. Take a look at the following examples.
```{r message = FALSE, warning = FALSE}
#use summarise to count observations in each continent
gapminder %>%
  filter(year %in% c(1952, 2007)) %>% #%in% identifies if an element belongs to a vector 
  group_by(continent) %>%
  summarise(n = n())

#the count function does both grouping and counting in a single function!
gapminder %>%
  filter(year %in% c(1952, 2007)) %>% 
  count(continent) 

#use summarise to count observations for each country-year and compute mean lifeExp
gapminder %>%
  filter(year %in% c(1952, 2007)) %>% 
  group_by(continent, year) %>%
  summarise(n = n(), avg_lifeExp = mean(lifeExp))

#let's use mutate to compute the change between 1952 and 2007 in country life expectancy
gapminder %>%
  filter(year %in% c(1952, 2007)) %>% 
  group_by(country) %>% 
  select(country, year, lifeExp) %>% 
  mutate(lifeExp_gain = lifeExp - first(lifeExp)) #create a new column using the grouped data

#here's a more advance example: obtaining extreme lifeExp values for each year
gapminder %>%
  select(year, country, lifeExp) %>%
  group_by(year) %>%
  filter(min_rank(desc(lifeExp)) < 2 | min_rank(lifeExp) < 2) %>% # min_rank is a dplyr ranking function
  arrange(year) %>%
  print(n = Inf) #Infinity can be used to override limits on the number of rows/columns printed
```


# Intro to ggplot for visualization

`ggplot2` is a package in the tidyverse that provides a system for creating data visualizations. It's a bit more difficult to learn than other R functionality because ggplot syntax relies on its own grammatical rules.

With `ggplot2`, you begin a plot with the function `ggplot()`. This creates a coordinate system that you can add layers to. We start with three components:

- **data**: the first argument of `ggplot()` is the data to use in the graph, e.g. `ggplot(data = gapminder)`. This creates an empty graph that you can add one or more layers to. 
- **aesthetics** map variables to visual properties like color, shape, size, etc. Aesthetic mappings are set as arguments to the `aes()` function, which must be specified along with a `ggplot()` call. `ggplot2` knows to look for mapped variables inside the data argument.
- **geometries** are geometric elements added to plots in layers. Examples includes points, lines, and bars (e.g. `geom_point()`, `geom_line()`, and `geom_bar()` or `geom_histogram()`).

Other components that you can specify include scale, statistical transformations, or facets that break information into subplots. [Here](https://ggplot2.tidyverse.org/reference/) is a handy reference sheet from ggplot2.

We'll cover ggplot data visualization in more detail in a few weeks, but for now let's review just the very basics needed to generate histograms and barplots as a tool for summarizing data.


## barplots

Let's learn the basic syntax by plotting the number of countries observed in each continent in the gapminder dataset for 2007. Barplots are generated using `geom_bar()` to plot categorical data, whereas histograms use `geom_histogram()` to visualize the distribution of a continuous variable.
```{r message = FALSE, warning = FALSE}
#filter gapminder to include 2007 observations only
gap_2007 <- gapminder %>% filter(year == 2007) 
```

```{r eval = FALSE}
#it can be helpful to include argument names while learning the syntax
ggplot(data = gap_2007) + aes(x = continent) + geom_bar()
```

```{r message = FALSE, warning = FALSE}
#but R is smart enough to recognize many arguments without specifying by name
ggplot(gap_2007) + aes(continent) + geom_bar()
```

Here is the same information with additional options to display continent proportions instead of counts.
```{r message = FALSE, warning = FALSE}
ggplot(gap_2007) + aes(x = continent, y = ..prop.., group = 1) + geom_bar()
```


`geom_bar()` works by simply counting observations in each category of the argument x (in this case `x = continent`). Alternatively, we can collapse the country-level data to continent level observations using `group_by()` and `summarise()`, and the plot the data directly using `geom_col()`. Don't worry too much about this for now, just note that the geometries we use in our plots depend on the structure of the data.
```{r message = FALSE, warning = FALSE}
gap_continents <- gapminder %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  summarise(country_ct = n())
ggplot(data = gap_continents) + aes(x = continent, y = country_ct) + geom_col()
```


## histograms

Here is the code to visualize the distribution of life expectancy across countries in 2007 using a very basic histogram with minimal additional options specified (just specified a bin width of 2 years as an optional argument).
```{r message = FALSE, warning = FALSE}
ggplot(gap_2007) + aes(x = lifeExp) + geom_histogram(binwidth = 2)
```


