---
title: 'Assignment 2 - Sample Solutions'
author: "Liam Tay Kearney"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
    toc_depth: '3'
    number_sections: yes
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: yes
    highlight: tango
    theme: default
    fig_caption: yes
    df_print: tibble
urlcolor: blue
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
```

##### *Please submit your knitted .pdf file along with the corresponding R markdown (.rmd) via Courseworks by 11:59pm on Monday, September 27th.* {-}

\medspace

*Before knitting your rmd file as a pdf, you will need to install TinyTex for Latex distribution by running the following code:*

`tinytex::install_tinytex()`

*Please visit [this](https://bookdown.org/yihui/rmarkdown-cookbook/install-latex.html) link for more information on TinyTex installation.*

\medspace

# Load and inspect CPS data: 

```{r}
cps <- read.csv("cps_june_20-21.csv")
cps <- na.omit(cps)
```

### a) Inspect the data frame and data types for each column {-}
- make sure to inspect the age, sex, race, college columns

```{r}
summary(cps$age)
summary(cps$sex)
summary(cps$race)
summary(cps$college)
```
### b) Use the mutate function to create new column for sex {-}
- sex.fac = as.factor(sex),
- check if it worked by calling the str() function

```{r eval=FALSE}
mutate(cps, sex.fac = as.factor(sex)) #output suppressed
```

```{r}
str(mutate(cps, sex.fac = as.factor(sex)))
```

### c) Include sex.fac in a new data frame called cps.temp1 {-}
- also create new factor columns for race and college education,
- in the same pipe, exclude the columns for serial and ind
- after creating cps.temp1, print the first 5 observations

```{r}
  cps.temp1 <- cps %>%
    mutate(sex.fac = as.factor(sex),
           race.fac = as.factor(race),
           college.fac = as.factor(college)) %>%
    select(-serial, -ind)
```

```{r}
  head(cps.temp1, n = 5)
```
```{r}
#A neater way to present (key data only, other cols omitted)
head(cps.temp1, n = 5) %>% 
  select(sex.fac, race.fac, college.fac, earnweek) %>% 
  knitr::kable()
```


### d) Inspect race.fac, sex.fac, and college.fac using the levels() function {-}
- what package is the levels() function located in?

```{r}
levels(cps.temp1$sex.fac)
levels(cps.temp1$race.fac)
levels(cps.temp1$college.fac)
#?levels #from the documentation, the levels function is located in base R.
```

### e) Use filter() to only include rows only for June 2020 {-}
- store as a new object cps_2020,
- print the first 5 observations,

```{r}
cps_2020 <- cps.temp1 %>% 
  filter(year == 2020)

head(cps_2020, n = 5)
```

- confirm your data only includes observations for 2020

```{r}
summary(cps_2020$year)
```

### f) Remove the cps.temp1 object from memory using the rm() function {-}

```{r}
rm(cps.temp1)
```

\newpage

# Describe the cps_2020 data frame

### a) What is the unit of observation? {-}

```{r}
str(cps_2020)
```
The unit of observation is the individual (individuals surveyed in June, 2020).


### b) How many individuals are observed? From how many households? {-}

```{r}
summarise(cps_2020, n_distinct(personid))
summarise(cps_2020, n_distinct(hhid))
```
```{r eval=FALSE}
#Alternative way:
cps_2020 %>%
  summarise(num_persons = n_distinct(personid),
            num_households = n_distinct(hhid))
```
There are `r n_distinct(cps_2020$personid)` unique individuals, and `r n_distinct(cps_2020$hhid)` unique households.

### c) What is the average age of individuals in the sample? Youngest and oldest person? {-}

```{r}
cps_2020 %>%
  summarise(avg_age = mean(age),
            min_age = min(age),
            max_age = max(age))
```

Alternatively, using inline code:
```{r}
sumstats <- cps_2020 %>%
              summarise(avg_age = mean(age),
                        min_age = min(age),
                        max_age = max(age))
```

The average age in the sample is `r round(sumstats$avg_age,2)`, with individuals ranging from `r sumstats$min_age` to `r sumstats$max_age` years old.

\newpage

# Earnings per week for different groups in June 2020

### a) Find the observation for the top weekly earnings using the summarise() function {-}
- assign this to a new object called max_earnings 

```{r}
max_earnings <- cps_2020 %>%
                  summarise(max_earnings = max(earnweek))
max_earnings
```

### b) Find max weekly earnings using the arrange function instead of summarise {-}
```{r}
cps_2020 %>% 
  arrange(desc(earnweek)) %>% 
  select(earnweek) %>% 
  head(n = 1)
```

### c) Use the filter function to subset for the observation with max weekly earnings {-}
- don't hardcode the max earnings to filter on, refer to the max_earnings object from a),
- store in new data frame cps_max_earn,

```{r}
cps_max_earn <- cps_2020 %>% 
                  filter(earnweek == max_earnings[1,])

#Alternative way:
cps_max_earn <- cps_2020 %>% 
                  arrange(desc(earnweek)) %>% 
                  head(n = 1)
```

- confirm it worked
```{r}
cps_max_earn %>% 
  select(sex, race, age, personid, college, earnweek)
```


### d) What is the age, sex, and race of the top weekly earner in the sample? {-}
```{r}
cps_max_earn %>%
  select(age,sex,race) %>%
  head(n = 1)
```
Alternatively, to make it look nicer, we can pipe the output to knitr::kable().
```{r}
cps_max_earn %>%
  select(age,sex,race) %>%
  head(n = 1) %>% 
  knitr::kable()
```


### e) List the age, sex, and race of the top 10 weekly earners in the sample {-}
```{r}
cps_2020 %>%
  arrange(desc(earnweek)) %>%
  select(age,sex,race, earnweek) %>% 
  head(n=10) %>% 
  knitr::kable()
```

### f) How many individuals earned more than $2000 in weekly earnings? {-}
```{r}
cps_2020 %>%
  filter(earnweek > 2000) %>%
  nrow()
```
\newpage


# Wage gaps between males and females:

### a) Use the filter function to subset observations for males {-}
- assign to new data frame, cps_2020_male,
- sort in descending order of weekly earnings
- check if it worked

```{r}
cps_2020_male <- cps_2020 %>%
                    filter(sex.fac == "Male") %>%
                    arrange(desc(earnweek))

#Check
cps_2020_male %>%
  select(sex.fac,earnweek) %>% 
  head(n = 3) %>% 
  knitr::kable()
```


### b) Repeat part a for females and create a new data frame, cps_2020_female {-}
```{r}
cps_2020_female <- cps_2020 %>%
                    filter(sex.fac == "Female") %>%
                    arrange(desc(earnweek))

#Check
cps_2020_female %>%
  select(sex.fac,earnweek) %>% 
  head(n = 3) %>% 
  knitr::kable()
```

\medspace

### c) Use summarise to find mean, min & max for males and females, separately {-}
- name each statistic appropriately (i.e. name each column in the 1-row table of stats)
```{r}
cps_2020_male %>%
  summarise(mean_earnings_male = mean(earnweek),
            min_earnings_male = min(earnweek),
            max_earnings_male = max(earnweek)) %>% 
  knitr::kable()

cps_2020_female %>%
  summarise(mean_earnings_female = mean(earnweek),
            min_earnings_female = min(earnweek),
            max_earnings_female = max(earnweek)) %>% 
  knitr::kable()
```

- what is the gender gap in mean weekly earnings?

The gender gap in weekly earnings is \$`r round(mean(cps_2020_male$earnweek) - mean(cps_2020_female$earnweek),2)`.

### d) What is the wage gap in weekly earnings between white males and Black females? {-}

```{r}
cps_2020_wh_male <- cps_2020_male %>%
                      filter(race.fac == "White")
```

```{r}
cps_2020_bl_female <- cps_2020_female %>%
                        filter(race.fac == "Black/Negro")
```

The weekly earnings gap between white males and Black females is \$`r round(mean(cps_2020_wh_male$earnweek) - mean(cps_2020_bl_female$earnweek),2)`.

### e) What is the wage gap between college educated white males and college educated Black females? {-}

```{r}
cps_2020_wh_male_college <- cps_2020_male %>%
                                filter(college.fac == "College degree" & 
                                         race.fac == "White")
```

```{r}
cps_2020_bl_female_college <- cps_2020_female %>%
                                filter(college.fac == "College degree" & 
                                         race.fac == "Black/Negro")
```

The weekly earnings gap between white college-educated males and Black college-educated females is \$`r round(mean(cps_2020_wh_male_college$earnweek) - mean(cps_2020_bl_female_college$earnweek),2)`.

*NOTE: the exercises above are done using weekly earnings, but can easily be converted to hourly wages*

\medspace

**End of assignment.**
