---
title: 'U6614: Assignment 2: COVID-19 Country Case Data'
author: "Your Name (your-uni)"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
    toc_depth: '3'
    number_sections: yes
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: yes
    highlight: tango
    theme: default
    fig_caption: yes
    df_print: tibble
urlcolor: blue
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

##### *Please submit your knitted .pdf file along with the corresponding R markdown (.rmd) via Courseworks by 11:59pm on Monday, January 25th.* {-}

\medspace

*Before knitting your rmd file as a pdf, you will need to install TinyTex for Latex distribution by running the following code:*

`tinytex::install_tinytex()`

*Please visit [this](https://bookdown.org/yihui/rmarkdown-cookbook/install-latex.html) link for more information on TinyTex installation.*

*If you run intro problems knitting as a pdf, please test by first knitting as an html file. If your rmd file knits as an html file but not as a pdf, then it is likely an RStudio/TinyTex installation issue. A quick fix is to open your knit html file in Chrome or your preferred browser, print as a pdf and submit that file to Courseworks.*

*If you are unable to knit as an html file, then the issue lies with your code and not your RStudio setup.*


# Introduction {-}

Load packages:
```{r}
library(tidyverse)
```


# Load and prep the data

#### Load the COVID-19 data from class and only keep observations without NA values for cases (not deaths). {-}

Data source: https://www.ecdc.europa.eu/en/publications-data/data-national-14-day-notification-rate-covid-19
\
\
```{r q1}
#load raw data
  covid <- read.csv("ecdc_national_2021-01-14.csv")
  covid <- na.omit(covid) #remove all observations with NA values

#clean raw data as needed and assign to new data frame to answer questions
#NOTE: this combines code from class into a single pipe
  covid_cases <- covid %>%
    mutate(ind.fac = as.factor(indicator)) %>%
    select(-indicator, -continent, -source) %>% 
    filter(ind.fac == "cases") %>% 
    mutate(ind.char = as.character(ind.fac),
           ind.fac = as.factor(ind.char)) %>%
    filter(country_code != "") %>% 
    select(country, year_week, rate_14_day, ind.fac)
  
  head(covid_cases, n = 10)
```
\


# Describe the data

#### Provide the following, along with any other information you think might be useful for the reader to know about the data. {-}

- *unit of observation*
- *date range observed in the data*
- *number of countries (or administrative entities reporting data)*

Each observation is the week-country pair; in other words, we observe every reporting country for each week in the period covered by the data.

```{r}
covid_dates <- covid_cases %>% 
  summarise(num_of_weeks = n_distinct(year_week), # num of countries
            firstweek = min(year_week),           #first week
            lastweek = max(year_week))            #last week
covid_dates
```

The data spans `r covid_dates$num_of_weeks` weeks, from `r covid_dates$firstweek` through `r covid_dates$lastweek`.


# Latest global 14-day case rate (per 100,000 population)

#### a. Create a new data frame that only includes observations for the most recent week. {-}

\medspace

*Note: don't hard-code a date to filter on, find the latest week, store as a data object, and then refer back to (the element in) that object (see Lecture2-inclass.r for guidance)*

```{r}
#find latest week
  lastweek <- covid_cases %>% summarise(max(year_week))

#filter on last week
  covid_cases_last <- covid_cases %>%      
    filter(year_week == lastweek[,1])
  head(covid_cases_last)
```


#### b. What was max 14-day rate for the most recent week observed in the data? {-}

\medspace
```{r}
#sort in descending order
  covid_cases_last_sort <- covid_cases_last %>% 
    arrange(desc(rate_14_day)) %>% 
    select(rate_14_day, country, year_week) 

#store first row in sorted data frame as its own object
  lastweek_max <- covid_cases_last_sort[1,]
```
The largest reported 14-day case rate per 100,000 population for `r lastweek_max[1,3]` was `r round(lastweek_max[1,1],0)` for `r lastweek_max[1,2]`.


#### c. List the top 10 countries by 14-day rate for the most recent week observed in the data? {-}

\medspace
```{r}
covid_cases_last %>% 
  select(rate_14_day, country) %>%
  arrange(desc(rate_14_day)) %>%
  head(n = 10)
```


#### d. How many countries had zero reported cases for the most recent week? {-}

\medspace
```{r}
zerocases <- covid_cases_last %>% 
  select(rate_14_day, country) %>%
  arrange(country) %>% 
  filter(rate_14_day == 0)
zerocases
```

There were `r nrow(zerocases)` countries reporting 0 COVID-19 cases for `r lastweek_max[1,2]`.


# Panama 14-day case rates

#### a.  Create a new data frame for the 14-day rate for Panama only. Sort in descending data order. {-}

\medspace
```{r}
covid_cases_panama <- covid_cases %>% 
  filter(country == "Panama") %>% 
  arrange(desc(year_week)) 

head(covid_cases_panama)
```


#### b.  Find the weekly mean, min, and max 14-day rate for Panama over all included data and name each column appropriately. {-}

\medspace
```{r}
panama <- covid_cases_panama %>% 
  summarise(panama_mean = mean(rate_14_day),
            panama_min = min(rate_14_day),
            panama_max = max(rate_14_day))
panama
```
The average 14-day case rate for Panama over the period covered in this data is `r round(panama$panama_mean,1)`, with daily counts ranging from `r round(panama$panama_min,1)` to `r round(panama$panama_max,1)`.



#### c.  What was the average 14-day rate in Panama over the *last* 10 weeks of reported data? {-}

\medspace

*[HINT: See Lecture2.1 -> Section 4.2 for examples of subsetting syntax that can help you refer to the first 10 rows of sorted data. If you're having trouble, you can also try using the `row_number()` function]*

```{r}
panama_last10weeks_mean <- covid_cases_panama %>%
  arrange(desc(year_week)) %>% 
  filter(row_number() <= 10) %>% 
  summarise(mean(rate_14_day))
```
The average 14-day rate in Panama over the *last* 10 weeks was `r round(panama_last10weeks_mean[1,],1)`.


#### d.  What was the average 14-day rate in Panama over the *first* 10 weeks of reported data? {-}

\medspace
```{r}
panama_first10weeks_mean <- covid_cases_panama %>%
  arrange(year_week) %>% 
  filter(row_number() <= 10) %>% 
  summarise(mean(rate_14_day))
```
The average 14-day rate in Panama over the *first* 10 weeks was `r round(panama_first10weeks_mean[1,],1)`.


