---
title: 'U6614: Assignment 3: Subway Fare Evasion Microdata'
author: "Sample Solution"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
    toc_depth: '3'
    number_sections: yes
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: yes
    highlight: tango
    theme: default
    fig_caption: yes
    df_print: tibble
urlcolor: blue
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

##### *Please submit your knitted .pdf file along with the corresponding R markdown (.rmd) via Courseworks by 11:59pm on Monday, February 1st.* {-}

\medspace

*Before knitting your rmd file as a pdf, you will need to install TinyTex for Latex distribution by running the following code:*

`tinytex::install_tinytex()`

*Please visit [this](https://bookdown.org/yihui/rmarkdown-cookbook/install-latex.html) link for more information on TinyTex installation.*

# Load libraries

```{r}
#remember to make sure these packaged are installed before trying to load
library(tidyverse)
library(fastDummies)
```


\medspace

# Load and inspect the two public defender client datasets (BDS & LAS).  

```{r}
arrests_bds <- read_csv("microdata_BDS_inclass.csv", na = "")
arrests_las <- read_csv("microdata_LAS_inclass.csv", na = "")

str(arrests_bds)
str(arrests_las)
```


## Give a brief overview of the data. The aim is not be exhaustive, but to paint a picture of they key features of the data with respect to the policy questions you'll be exploring.

The BDS data includes `r nrow(arrests_bds)` observations (client arrest records), and the LAS data includes another `r nrow(arrests_las)` observations. Both datasets include basic demographic information on age, sex, race, ethnicity (coded differently in each dataset), as well as information on the location/subway station where the arrest occurred. 

The LAS data also includes information on case dismissal rates.


## For each dataset, what is the unit of observation and population represented by this "sample"? Do you think this sample does a good job representing the population of interest?

In each raw dataset, the unit of observation is the arrested individual (client). On the surface, the representative population is all individuals arrested by the NYPD for subway fare evasion in Brooklyn during 2016 who are represented by public defenders. If nearly all individuals arrested for fare evasion are represented by public defenders, then this sample comes close to the universe of subway fare evasion arrests in Brooklyn in 2016. This is difficult to argue convincingly without additional information, but is supported anecdotally by court observers.


## Inspect and describe the coding of race/ethnicity in each dataset.

```{r}
#recode race/ethnicity information from character to factors
arrests_bds <- arrests_bds %>% mutate(race = as.factor(race), 
                                      ethnicity = as.factor(ethnicity) )
arrests_las <- arrests_las %>% mutate(race = as.factor(las_race_key), 
                                      ethnicity = as.factor(hispanic_flag) )
#compare race coding
  summary(arrests_bds$race)
  summary(arrests_las$race)

#compare Hispanic/ethnicity coding
  summary(arrests_bds$ethnicity)
  summary(arrests_las$ethnicity)
```

Race information is generally stored in one variable, Hispanic identity in a second variable. To work towards consistent variable names and coding in both datasets, let's first recode the raw race and ethnicity information into two separate columns of data (factors) named `race` and `ethnicity`.

While each dataset refers to similar race and ethnicity categories, there are different category names in each (including some slightly different spellings). 

We also note that Hispanic identity factors into both race and Hispanic variables in the Legal Aid Society (LAS) data; in the BDS data, information on Hispanic identity is only included in the ethnicity variable.

Each dataset also contains a different set of values that seem to convey unknown race/ethnicity information, in addition to true missings (e.g. "0" and "Unknown" in addition to blank entries).


## From the outset, are there any data limitations you think are important to note?

It's unclear what processes are used to code race and ethnicity at each public defender group. How much does the information reflect client self-identification rather than identity assigned by police and entered into arrest reports?

It's also important to emphasize what information this data does **not** include that might be relevant to the question of biased fare evasion enforcement:

- fare evasion that resulted in a summons (ticket + fine) rather than an arrest
- fare evasion enforcement on buses


\medspace

# Clean BDS race and ethnicity data (insert code chunks that only include code you used to recode and very briefly validate your recoding )


## BDS: race data (generate column `race_clean`).

```{r}
#identify every combination of race-ethnicity in the raw data
table(arrests_bds$race, arrests_bds$ethnicity, useNA = "always")

#recode as factor in an internally consistent manner (address NAs, specify levels)
arrests_bds.clean <- arrests_bds %>% 
  mutate(race_clean = recode(race, "0" = "NA", 
                                   "Unknown" = "NA", 
                                   "Am Indian" = "Other" ) )  %>% 
  mutate(race_clean = factor(race_clean, 
                             levels = c("Black", "White", "Asian/Pacific Islander", "Other")))

#validation: confirm the recode worked as intended
arrests_bds.clean %>% count(race_clean, sort = TRUE)
table(arrests_bds.clean$race_clean, arrests_bds.clean$race, useNA = "always")
```


## BDS: ethnicity data (generate column `ethnicity_clean`).

```{r}
#ok now let's recode to Hispanic, Non-Hispanic, and NA
arrests_bds.clean <- arrests_bds.clean %>% 
  mutate(hispanic = recode(ethnicity, "0" = "NA", "Other" = "Non-Hispanic") ) %>%
  mutate(hispanic = factor(hispanic, levels = c("Hispanic", "Non-Hispanic"))) 

#validation: confirm the recode worked as intended
  summary(arrests_bds.clean$hispanic)
  table(arrests_bds.clean$race_clean, arrests_bds.clean$hispanic, useNA = "always")
```


## Generate a single race/ethnicity factor variable `race_eth` with mutually exclusive categories.

```{r}
#let's investigate a bit
table(arrests_bds.clean$race_clean, arrests_bds.clean$hispanic, useNA = "always")

#generate race_eth column (as a factor) in steps
arrests_bds.clean <- arrests_bds.clean %>% 
  mutate(race_clean_char = as.character(race_clean)) %>% #work with characters
  mutate(hispanic_char = as.character(hispanic))     %>% #work with characters
  mutate(race_eth = ifelse(hispanic_char == "Hispanic", 
                           hispanic_char, 
                           race_clean_char) ) %>%  
  mutate(race_eth = as.factor(recode(race_eth, "White" = "Non-Hispanic White"))) %>%
  select(-race_clean_char, -hispanic_char)

#validate results: joint distribution of race_eth and hispanic
table(arrests_bds.clean$race_eth, arrests_bds.clean$hispanic, useNA = "always")
arrests_bds.clean %>% count(race_eth, sort = TRUE)
```

Note that `race_eth` assigns individuals who identify as both Hispanic and a race other than white as Hispanic. This means, for example, that an individual who identifies as both Black and Hispanic appears as Hispanic in the `race_eth` column.


\medspace

# Clean LAS race and ethnicity data

## Follow your own steps to end up at a `race_eth` variable for the LAS data that is coded in a comparable manner as in the BDS data.

*NOTE: you may be able to do everything in a single pipe, depending on your approach (but you certainly don't have to).*

```{r}
#inspect LAS data
table(arrests_las$las_race_key, arrests_las$hispanic_flag, useNA = "always")

#generate race_eth column as a factor
arrests_las.clean <- arrests_las %>% 
  mutate(race_eth = recode(las_race_key, "Asian or Pacific Islander" = "Asian/Pacific Islander",
                                         "Unknown" = "NA",
                                         "Latino" = "Hispanic",
                                         "White" = "Non-Hispanic White")) %>% 
  mutate(race_eth = ifelse(hispanic_flag == "Y", "Hispanic", race_eth) ) %>% 
  mutate(race_eth = factor(race_eth, 
                           levels = c("Black", "Hispanic", "Non-Hispanic White", "Asian/Pacific Islander", "Other")))

#validate
arrests_las.clean %>% count(race_eth, sort = TRUE)
table(arrests_las.clean$race_eth, arrests_las.clean$hispanic_flag, useNA = "always")


```


\medspace

# Combining (appending) the BDS and LAS microdata 

## Create a column (`pd`) to identify public defender data source.

```{r}
arrests_bds.clean <- arrests_bds.clean %>% mutate(pd = "bds")
arrests_las.clean <- arrests_las.clean %>% mutate(pd = "las")
```


## Append `arrests_bds.clean` and `arrests_las.clean` using `rbind()`. Store as new data frame `arrests_all` and inspect for consistency/accuracy.

```{r}
  arrests.clean <- plyr::rbind.fill(arrests_las.clean, arrests_bds.clean) %>%
    mutate(pd = as.factor(pd),
           st_id = as.factor(st_id),
           loc2 = as.factor(loc2)) %>% #station/location info is not continuous
    select(pd, race_eth, age, male, dismissal, st_id, loc2)
  summary(arrests.clean)
```


## What is the total number of subway fare evasion arrest records?

The total number of subway fare evasion arrest records from both BDS and LAS is `r nrow(arrests.clean)`.


## Export `arrests_all` as .csv, and save as .rds file.

```{r}
write_csv(arrests.clean, "arrests_all.csv") 
saveRDS(arrests.clean, "../Lecture4/arrests.clean.rds")
```


\medspace

# Descriptive statistics by race/ethnicity

## Print the number of arrests for each race/ethnicity category (a frequency table).

```{r}
arrests.clean %>% count(race_eth, sort = TRUE)
```


## Print the proportion of total arrests for each race/ethnicity category.

```{r}
#including NAs
prop.table(table(arrests.clean$race_eth, useNA = "always")) %>% 
  round(2) %>% 
  as.data.frame() %>% 
  arrange(desc(Freq)) %>% 
  rename(race_eth = Var1)

#excluding NAs
prop.table(table(arrests.clean$race_eth)) %>% 
  round(2) %>% 
  as.data.frame() %>% 
  arrange(desc(Freq)) %>% 
  rename(race_eth = Var1)
```


## Show the average age, share male, and dimissal rate for each race/ethnicity category. Describe any noteworthy findings.

```{r}
arrests.clean %>%
  group_by(race_eth) %>%
  summarise(n = n(), 
            mean_age = mean(age, na.rm = TRUE),
            mean_male = mean(male, na.rm = TRUE),
            mean_dism = mean(dismissal, na.rm = TRUE))
```


\medspace

# Subway-station level analysis

## Create dummy variables for each race/ethnicity category and show summary statistics only for these dummy variables.

```{r}
arrests.clean <- dummy_cols(arrests.clean, select_columns = "race_eth")
arrests.clean %>% 
  summarise(mean_black = round(mean(race_eth_Black, na.rm = TRUE), 2),
            mean_hisp = round(mean(race_eth_Hispanic, na.rm = TRUE), 2),
            mean_nhw  = round(mean(`race_eth_Non-Hispanic White`, na.rm = TRUE), 2),
            mean_api  = round(mean(`race_eth_Asian/Pacific Islander`, na.rm = TRUE), 2),
            mean_oth  = round(mean(race_eth_Other, na.rm = TRUE), 2),
            mean_NA   = round(mean(race_eth_NA, na.rm = TRUE), 2)  )
```


## Aggregate to station-level observations and show a table with the top 10 stations by arrest totals, including the following information for each station:
 
  - *station name (given by loc2)*
  - *st_id*
  - *total number of arrests at each station*
  - *total number of arrests for each race_eth category at each station*
  - *sort in descending order of total number of arrests*
  - *remember to only show the top 10 stations*
  - *use kable() in the knitr package for better formatting*
  
```{r}
arrests_stations <- arrests.clean %>%  
  group_by(loc2) %>%
  summarise(st_id = first(st_id), 
            n = n(),
            n_black = sum(race_eth_Black, na.rm = TRUE), 
            n_hisp  = sum(race_eth_Hispanic, na.rm = TRUE),
            n_api   = sum(`race_eth_Asian/Pacific Islander`, na.rm = TRUE),
            n_nhw   = sum(`race_eth_Non-Hispanic White`, na.rm = TRUE), 
            n_oth   = sum(race_eth_Other, na.rm = TRUE) )   %>%
  arrange(desc(n))
knitr::kable(head(arrests_stations, n = 10)) 
```
  

##  Aggregate to station-level observations (group by loc2), and show a table of stations with at least 50 arrests along with the following information:

  - *station name (loc2)*
  - *station arrest total*
  - *combined total number of Black and Hispanic arrests*
  - *total number of arrests with race/ethnicity coded as NA*
  - *share of arrests that are Black and Hispanic (excluding race_eth = NA from denominator)*
  - *sorted in ascending order above Black and Hispanic arrest share*
  - *remember to only show stations with at least 50 total arrests*
  - *use kable() in the knitr package for better formatting*

```{r}
arrests_stations_top <- arrests.clean %>%  
  group_by(loc2)    %>%
  summarise(st_id = first(st_id), 
            n = n(),
            n_black = sum(race_eth_Black, na.rm = TRUE), 
            n_hisp  = sum(race_eth_Hispanic, na.rm = TRUE),
            n_api   = sum(`race_eth_Asian/Pacific Islander`, na.rm = TRUE),
            n_nhw   = sum(`race_eth_Non-Hispanic White`, na.rm = TRUE), 
            n_oth   = sum(race_eth_Other, na.rm = TRUE), 
            n_bh    = sum(race_eth_Black, race_eth_Hispanic, na.rm = TRUE),
            n_na    = sum(race_eth_NA))  %>%
  mutate(sh_bh = round(n_bh / (n - n_na), 2)) %>% 
  select(loc2, n, n_bh, n_na, sh_bh) %>% 
  filter(n >= 50) %>% 
  arrange(sh_bh)
knitr::kable(arrests_stations_top)
```


## Briefly summarize any noteworthy findings from the table you just generated.

At every single high-arrest subway station, the majority of arrested individuals are Black or Hispanic. This isn't surprising, given that 87 percent of *all* arrested individuals with coded race/ethnicity are Black or Hispanic.


# (OPTIONAL) Visualize the distribution of arrests by race/ethnicity at stations with > 100 arrests.

|        *Hint: see R code from class, section 8*

```{r}
#get data frame with obs for every station-race_eth pairings on arrest counts
arrests_stations_race <- arrests.clean %>%  
  group_by(loc2) %>% 
  mutate(st_arrests = n()) %>% 
  ungroup() %>% 
  group_by(loc2, race_eth)  %>%
  summarise(arrests = n(), st_arrests = first(st_arrests)) %>% 
  arrange(desc(st_arrests)) %>% 
  filter(st_arrests > 100)
arrests_stations_race

ggplot(arrests_stations_race, 
       aes(x = reorder(loc2, -st_arrests), y = arrests, fill = race_eth)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```



