---
title: 'U6614: Assignment 2: COVID-19 Country Case Data'
author: "Your Name (your-uni)"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
    toc_depth: '3'
    number_sections: yes
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: yes
    highlight: tango
    theme: default
    fig_caption: yes
    df_print: tibble
urlcolor: blue
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

##### *Please submit your knitted .pdf file along with the corresponding R markdown (.rmd) via Courseworks by 11:59pm on Monday, January 25th.* {-}

\medspace

*Before knitting your rmd file as a pdf, you will need to install TinyTex for Latex distribution by running the following code:*

`tinytex::install_tinytex()`

*Please visit [this](https://bookdown.org/yihui/rmarkdown-cookbook/install-latex.html) link for more information on TinyTex installation.*

*If you run intro problems knitting as a pdf, please test by first knitting as an html file. If your rmd file knits as an html file but not as a pdf, then it is likely an RStudio/TinyTex installation issue. A quick fix is to open your knit html file in Chrome or your preferred browser, print as a pdf and submit that file to Courseworks.*

*If you are unable to knit as an html file, then the issue lies with your code and not your RStudio setup.*


# Introduction {-}

Load packages:
```{r}
#load packages here
```


# Load and prep the data

#### Load the COVID-19 data from class and only keep observations without NA values for cases (not deaths). {-}

Data source: https://www.ecdc.europa.eu/en/publications-data/data-national-14-day-notification-rate-covid-19
\
\
[FOR EACH QUESTION, ADD YOUR WRITE-UP (IF APPLICABLE) UNDER THE QUESTIONS/HEADERS FOR ORGANIZATION]
\
```{r q1}
#include any code you use to arrive at your answers as code chunks
#but don't include code you don't need to arrive at your answers!
#remember to use comments liberally to explain/organize your code
```
In your written answers, use inline code chunks to refer to any numbers from your analysis rather than hard-coding numbers. See Lecture2.2, Section 5.3 for an overview of in-line code chunks.


# Describe the data

#### Provide the following, along with any other information you think might be useful for the reader to know about the data. {-}

- *unit of observation*
- *date range observed in the data*
- *number of countries (or administrative entities reporting data)*


# Latest global 14-day case rate (per 100,000 population)

#### a. Create a new data frame that only includes observations for the most recent week. {-}

\medspace

*Note: don't hard-code a date to filter on, find the latest week, store as a data object, and then refer back to (the element in) that object (see Lecture2-inclass.r for guidance)*



#### b. What was the max 14-day rate for the most recent week observed in the data? {-}


#### c. List the top 10 countries by 14-day rate for the most recent week observed in the data? {-}


#### d. How many countries had zero reported cases for the most recent week? {-}


# Panama 14-day case rates

#### a.  Create a new data frame for the 14-day rate for Panama only. Sort in descending data order. {-}


#### b.  Find the weekly mean, min, and max 14-day rate for Panama over all included data and name each column appropriately. {-}


#### c.  What was the average 14-day rate in Panama over the *last* 10 weeks of reported data? {-}

\medspace

*[HINT: See Lecture2.1 -> Section 4.2 for examples of subsetting syntax that can help you refer to the first 10 rows of sorted data. If you're having trouble, you can also try using the `row_number()` function]*


#### d.  What was the average 14-day rate in Panama over the *first* 10 weeks of reported data? {-}


