---
title: "U6614: Assignment 2: COVID-19 Country Case Data"
author: "SAMPLE SOLUTION"
date: "`r Sys.Date()`"
urlcolor: blue
output: 
  pdf_document:
    toc: no
    toc_depth: '3'
    number_sections: yes
  html_document:
    #fig_width = 7 #Default width (in inches) for figures
    #fig_height = 5 #Default height (in inches) for figures
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: yes
    highlight: tango
    theme: default
    fig_caption: yes
    #html_preview: TRUE    
    df_print: tibble # Method to be used for printing data frames. Valid values include "default", "kable", "tibble", and "paged". The "default" method uses print.data.frame. 
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

##### *Please submit your knitted .pdf file along with the corresponding R markdown (.rmd) via Courseworks by 11:59pm on Monday, September 21st.* {-}

# Introduction. {-}

Load packages:
```{r}
library(tidyverse)
```


# Load and prep the data.

#### Load the coronavirus.rda data from class and only keep confirmed cases. {-}

Data source: https://github.com/RamiKrispin/coronavirus/tree/master/data
\
\
```{r}
#load raw data
load("coronavirus.rda")

#inspect raw data
str(coronavirus) 

#clean raw data as needed and assign to new data frame
covid_confirmed.df <- coronavirus %>% 
  mutate(coronavirus, type.fac = as.factor(type)) %>% #convert type to factor
  filter(type.fac == "confirmed") %>%                 #keep confirmed cases
  mutate(country = if_else(province == "", 
                           country, 
                           paste(country, "-", province))) %>% #use province
  select(-type, -province) #drop unused type and province columns

head(covid_confirmed.df)
```
\

# Describe the data.

#### Provide the following, along with any other information you think might be useful for the reader to know about the data. {-}

- *unit of observation*
- *date range observed in the data*
- *number of countries (or administrative entities reporting data)*

The unit of observation is day-country; more accurately, observations are for countries except in select cases when the reporting entity is at the sub-country level such as a province (e.g. Xinjiang) or autonomous territory (e.g. Aruba). 

```{r}
covid_dates <- covid_confirmed.df %>% 
  summarise(num_of_days = n_distinct(date), #reporting entities
            firstday = min(date), #first date
            lastday = max(date))  #last date
covid_dates
```
The data spans `r covid_dates$num_of_days` days, from `r covid_dates$firstday` through `r covid_dates$lastday`.


```{r}
num_of_countries <- covid_confirmed.df %>% 
  summarise(num_of_countries = n_distinct(country))
num_of_countries
```
The data includes case counts for `r num_of_countries[1,]` different reporting entities. 


# Latest global case counts.

#### a. Create a new data frame that only includes observations for the most recent day only. {-}

*Note: don't hard-code a date to filter on, find the last day, store as a data object, and then refer back to (the element in) that object (see Lecture2-inclass.r for guidance)*
\
```{r}
lastday <- covid_confirmed.df %>% summarise(max(date)) #find last date
covid_confirmed_last.df <- covid_confirmed.df %>%      #filter on last date
  filter(date == lastday[,1])
```


#### b. What was max case count for the most recent day observed in the data? {-}

```{r}
lastday_max <- covid_confirmed_last.df %>% summarise(max(cases))
lastday_max

#short way
max(covid_confirmed_last.df$cases)
```
The largest reported case count for September 9th was `r lastday_max` for India.


#### c. List the top 5 countries (or administrative entities) by case count for the most recent day observed in the data? {-}

```{r}
covid_confirmed_last.df %>% 
  select(country, cases) %>%
  arrange(desc(cases)) %>%
  head(n = 5)
```


#### d. How many countries (or administrative entities) had zero confirmed cases for the most recent day? {-}

```{r}
zerocases <- covid_confirmed_last.df %>% 
  select(country, cases) %>%
  arrange(country) %>% 
  filter(cases == 0)
zerocases
nrow(zerocases)
```

There were `r nrow(zerocases)` countries reporting 0 COVID-19 cases on September 9th.



# Oman case counts.

#### a.  Create a new data frame for daily confirmed case counts for Oman only. Sort in descending data order. {-}

```{r}
covid_confirmed_oman.df <- covid_confirmed.df %>% 
  filter(country == "Oman") %>% 
  arrange(desc(date))
head(covid_confirmed_oman.df)
```


#### b.  Find the daily mean, min, and max case counts for Oman over the duration of the pandemic and name each column appropriately. {-}

```{r}
oman <- covid_confirmed_oman.df %>% 
  summarise(oman_mean = mean(cases),
            oman_min = min(cases),
            oman_max = max(cases))
oman
```
The average daily case count for Oman over the period covered in this data is `r oman$oman_mean`, with daily counts ranging from `r oman$oman_min` to `r oman$oman_max`.


#### c.  What was the average daily case count in Oman over *last* 30 days of reported data? {-}

*[HINT: See Lecture2.1 -> Section 4.2 for examples of subsetting syntax that can help you refer to the first 30 rows of sorted data. If you're having trouble, you can also try using the `filter()` function and use the `row_number()` function as a part of the filtering condition]*

```{r}
oman_avg_last30 <- covid_confirmed_oman.df[1:30,] %>% 
  summarise(oman_last30days_mean = mean(cases))
oman_avg_last30
```
The average daily case count in Oman over the last 30 days of reported data was `r oman_avg_last30[1,]`.


#### d.  What was the average daily case count in Oman over the *first* 30 days of reported data?{-}

```{r}
oman_avg_first30 <- covid_confirmed_oman.df %>% 
  arrange(date) %>% 
  filter(row_number() < 30) %>% 
  summarise(oman_first30days_mean = mean(cases))
oman_avg_first30
```
The average daily case count in Oman over the *first* 30 days of reported data is `r oman_avg_first30[1,]`.

